\Large{Learning to Pivot with Adversarial Networks}

### 2016-10-31 tex/paper.tex
add axis labels

In this section, we show that in the setting of
Algorithm~\ref{alg:adversarial-training} where ${\cal L}_f$ and ${\cal L}_r$ are
respectively set to expected value of the negative log-likelihood of $Y|X$ under
$f$ and of $Z|f(X;\theta_f)$ under $r$, the procedure converges to a classifier
$f$ which is a pivotal quantity in the sense of Eqn.~\ref{eqn:criterion}.

@@

Assuming a gaussian prior $z \sim {\cal N}(0,1)$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we train a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without
considering its adversary $r$. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in the upper plots of
Fig.~\ref{fig:toy}, the resulting classifier is not pivotal, as the
conditional probability densities of its decision scores $f(X;\theta_f)$ show
large discrepancies between values $z$ of the nuisance. While not shown here, a
classifier trained only from data generated at the nominal value $Z=0$ would
also not be pivotal.

Let us now consider the joint training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, output nodes for the standard deviations with exponential
activations to ensure positivity, while output nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Fig.~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Fig.~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in the third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $H({Y|X}) -
\lambda H(Z)$.

@@

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/ams.pdf}
    \vspace{-1cm}
    \caption{Physics example. Approximate median significance as a function of the decision threshold
             on the output of $f$.  As shown at $\lambda=10$, trading
             classification accuracy for independence with respect to pileup
             results in a positive total net benefit in terms of statistical significance.}
    \label{fig:physics-ams}
\end{figure}

>>>>

In this section, we show that in the setting of
Algorithm~\ref{alg:adversarial-training} where ${\cal L}_f$ and ${\cal L}_r$ are
respectively set to expected value of the negative log-likelihood of $Y|X$ under
$f$ and of $Z|f(X;\theta_f)$ under $r$, the minimax solution of Eqn.~\ref{eqn:min_thetaf}
corresponds to a classifier
$f$ which is a pivotal quantity.

@@

Assuming a gaussian prior $z \sim {\cal N}(0,1)$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we train a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without
considering its adversary $r$. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in 
Fig.~\ref{fig:toy}, the resulting classifier is not pivotal, as the
conditional probability densities of its decision scores $f(X;\theta_f)$ show
large discrepancies between values $z$ of the nuisance. While not shown here, a
classifier trained only from data generated at the nominal value $Z=0$ would
also not be pivotal.

Let us now consider the joint training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, output nodes for the standard deviations with exponential
activations to ensure positivity, while output nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in Fig.~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Fig.~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in the third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $H({Y|X}) -
\lambda H(Z)$.

@@

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/ams.pdf}
    \caption{Physics example. Approximate median significance as a function of the decision threshold
             on the output of $f$.  As shown at $\lambda=10$, trading
             classification accuracy for independence with respect to pileup
             results in a positive total net benefit in terms of statistical significance.}
    \label{fig:physics-ams}
\end{figure}
